# Activity Monitoring Analysis
## Reproducible Research - Project 1
author: Roberto Garuti

```{r setoption, echo=TRUE}
```
Load the packages needed for the analysis and the data from the working directory
```{r load, message=FALSE}
library("dplyr", lib.loc="~/R/win-library/3.3")
library("lattice", lib.loc="C:/Program Files/R/R-3.3.2/library")
stepdata <- read.csv("activity.csv", header = TRUE)
```
Calculate the total number of steps recorded in each day and plot it on an histogram
```{r dailysteps}
dailysteps <- aggregate(steps~date, stepdata, sum )
hist(dailysteps$steps, main = "Total Steps by Day", xlab = "Steps per day", breaks = 10, 
     col = "blue")
```

Calculate the mean value for the steps recorded in each day
```{r dailymean}
dailymean <- mean(dailysteps$steps, na.rm = TRUE)
dailymean
```
Calculate the median value for the steps recorded in each day
```{r dailymedian}
dailymedian <- median(dailysteps$steps, na.rm = TRUE)
dailymedian
```
Calculate the mean value for the number of steps taken in each 5-minute interval 
averaged across each day and plot the result in a time series graph
```{r 5minavgsteps}
intervalsteps <- aggregate(steps~interval, stepdata, mean )
plot(intervalsteps$interval, intervalsteps$steps, type = "l", xlab = "Interval", 
     ylab = "Average Steps", main = "Average Steps for Each 5-minute Interval")
```

The interval corresponding to 8:35 is the one with the highest mean number of 
steps across all days (206.1698 in 5 minutes)
```{r max5minavg}
intervalsteps[intervalsteps$steps == max(intervalsteps$steps),]
```
Calculate total number of missing values in the dataset (NAs)
```{r NAstot}
length(stepdata$steps[is.na(stepdata$steps)])
```
Replace each missing value in the dataset with the corresponding average number of 
steps for that 5-minute interval calculated across all days in the previous step
```{r replaceNAs}
for(i in 1:nrow(stepdata)) {
  if(is.na(stepdata[i,1])) {
    y <- stepdata[i,3]
    z <- intervalsteps[intervalsteps$interval == y,2]
    stepdata[i, 1] <- z
  }
}
```
Calculate the total number of steps recorded in each day and plot it on an histogram
for this new dataset with filled NAs
```{r r dailystepsnoNAs}
dailysteps1 <- aggregate(steps~date, stepdata, sum )
hist(dailysteps1$steps, main = "Total Steps by Day", xlab = "Steps per day", breaks = 10, 
     col = "blue")
```

Calculate the mean value for the steps recorded in each day for this new dataset.
This value is identical to the one calculated for the original dataset. Filling the NAs
with the selected strategy didn't affect the mean value since the NAs were filled with
daily means themselves
```{r dailymeannoNAs}
dailymean1 <- mean(dailysteps1$steps, na.rm = TRUE)
dailymean1
```
The same process is performed for the median value. In this case the value slightly
changes especially because the imputed values, being averages, are not necessarily
integers while the original median is calculated based on recorded data that have
integer values for the number of steps in each 5 minute interval.
```{r dailymediannoNAs}
dailymedian1 <- median(dailysteps1$steps, na.rm = TRUE)
dailymedian1
```
Add a factor to the dataset indicating if the day is a weekday or a weekend day
```{r weekdayfactor}
stepdata$date <- as.Date(stepdata$date,"%Y-%m-%d" )
stepdata$weekday <- weekdays(stepdata$date)
stepdata$weekday <- as.factor(stepdata$weekday)
stepdata[stepdata$weekday %in% c("Saturday","Sunday"),5] <- "weekend"
stepdata[!(stepdata$weekday %in% c("Saturday","Sunday")),5] <- "weekday"
names(stepdata)[5] <- "weekworkday"
stepdata$weekworkday <- as.factor(stepdata$weekworkday)
```
Create a panel plot for the mean value of the number of steps taken in each 5-minute interval averaged across each day, distinguishing between weekdays and weekend days
```{r weekdayplot}
wkdataframe <- aggregate(steps~interval+weekworkday,stepdata,mean)
xyplot(steps~interval|weekworkday,data = wkdataframe,layout=c(1,2), type="l",
       ylab = "Average Steps")
```